Local Spectral Clustering of Density Upper Level Sets

### 2019-11-18 arxiv/local_spectral_density_clustering_arxiv.tex
Made my pass through the paper

\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem*{remark}{Remark}

@@

\newcommand{\diam}{\rho}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\seq}[1]{\left\{#1\right\}_{n \in \mathbb{N}}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\Phi_{u}}
\newcommand{\Psibf}{\Psi_{u}}
\newcommand{\taubf}{\tau_{u}}
\newcommand{\dist}{\mathrm{dist}}
\newcommand{\Err}{\mathrm{Err}}
\newcommand{\TV}{\mathrm{TV}}

%%% Vectors
\newcommand{\pbf}{p}        % removed bold font
\newcommand{\qbf}{\mathbf{q}}
\newcommand{\ebf}[1]{{e}_{#1}}
\newcommand{\pibf}{\pi}

%%% Matrices (no bold font)
\newcommand{\Abf}{A}
\newcommand{\Xbf}{X}             % removed bold font 
\newcommand{\Wbf}{W}
\newcommand{\Lbf}{L}
\newcommand{\Dbf}{D}
\newcommand{\Ibf}[1]{I_{#1}}

@@

{\large{
\begin{center}
Alden Green~~~~~ Sivaraman Balakrishnan~~~~~ Ryan Tibshirani\\
\vspace{.2cm}
\end{center}

@@

\begin{tabular}{c}
\texttt{\{ajgreen,sbalakri,ryantibs\}@andrew.cmu.edu}
\end{tabular}
}}

@@

\begin{abstract}
\noindent We analyze the Personalized PageRank (PPR) algorithm, a local spectral method for clustering,
which extracts clusters using locally-biased random walks around a user-specified
seed node.  In contrast to previous work, we adopt a classical statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also show a surprising converse result, that PPR can fail 
to recover geometrically poorly conditioned density clusters, even asymptotically. Finally, we provide empirical support to illustrate our theoretical results.
\end{abstract}
\end{center}
\section{Introduction}
\label{sec: introduction}
In this paper, we consider the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  Our particular focus is on
spectral clustering methods, a family of powerful
nonparametric clustering algorithms. Roughly speaking, a spectral algorithm
first constructs a geometric graph $G$, where vertices correspond to samples,
and edges correspond to proximities between samples. The algorithm 
then estimates a feature
embedding based on (an appropriate) Laplacian matrix of the graph 
$G$, and applies a simple clustering
technique (like $k$-means clustering) in the embedded feature~space.

% Spectral clustering and density clustering
%Spectral clustering algorithms, which
%originated in the early work of Fiedler \cite{fiedler73}, have seen a recent revival 
%(see for instance the recent review~\cite{XX}) focused mainly on understanding the performance of spectral clustering on the Stochastic Block Model (SBM) and closely related problems.
%A parallel line of work \cite{shi09,schiebinger15}, has considered spectral partitioning applied to a non-parametric mixture model
%where once again the clustering goal is to recover the latent labels of a sample $X_1,\ldots,X_n \sim P$. These works proceed by relating the eigenvectors of the graph Laplacian to eigenfunctions of an appropriately defined integral operator, and then using assumptions about the (non-)overlap of the mixture components to show that these eigenfunctions elucidate the latent labels.

When applied to geometric graphs built from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to increased interest in
\emph{local} spectral clustering algorithms, which leverage locally-biased spectra
computed using random walks around some user-specified seed node.  A popular 
local clustering algorithm is the Personalized PageRank (PPR) algorithm, first introduced by 
\citet{haveliwala2003}, then further developed by
several others~\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013}.

Local spectral clustering techniques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, leading 
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (such as
conductance).  In this paper, we adopt a classical statistical viewpoint,
and examine what the output of local clustering on a data set reveals about the
underlying density $f$ of the samples.  In particular, we examine the ability of PPR to recover
\emph{density clusters} of $f$, defined as the connected components of
the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$ for some $\lambda > 0$
(a central object of interest in the statistical clustering literature, dating
back to the work of \citet{hartigan1981}).

\section{Background and Related Work}
We begin by providing some standard background on the PPR algorithm
and the density clustering setup, before turning our attention to related work
and a
detailed summary of our contributions.

\subsection{PPR on a neighborhood graph} 
We now describe the clustering
algorithm that will be our focus for the rest of the paper. Let $\Xbf = \{x_1,
\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$,
with density~$f$.  For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the
\emph{$r$-neighborhood graph} of $\Xbf$, an unweighted, undirected graph with
vertices $V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i -
x_j} \leq r$, where $\norm{\cdot}$ is the Euclidean norm. We denote by $\Abf \in
\Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if
$(u,v) \in E$ and $0$ otherwise.  We also denote by $\Dbf$ the diagonal degree
matrix, with entries $\Dbf_{uu} := \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n
\times n$ identity matrix.

Next, we define the PPR vector $\pbf_v = \pbf(v,\alpha;G_{n,r})$, based on
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be 
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf_v = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in the
$v$-th position and 0 elsewhere).

\subsubsection{Sweep cut}
\label{subsection: sweep_cut}
We employ the sweep cut method to obtain a local cluster from the PPR embedding $\pbf_v$. In more detail, for a level $\beta > 0$, we define a \emph{$\beta$-sweep cut} of $\pbf_v = (\pbf_v(u))_{u \in V}$
as   
\begin{equation}
\label{eqn: sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{\pbf_v(u)}{\Dbf_{uu}} > \beta}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(L, U)$} (where the range $(L,U)$ is also a user-specified parameter), we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} that has minimum normalized cut.
%\smash{$\Phi(S_{\beta^*}; G_{n,r})$}. 
For concreteness, the PPR algorithm is summarized in Algorithm~\ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, sweep cut range $(L,U)$. \\     
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf_v=\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn: 
    ppr_vector}. 
  \STATE For \smash{$\beta \in (L,U)$} compute sweep cuts 
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return the cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (L,U)}~ \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Estimation of density clusters} 
Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The size of the symmetric set difference between estimated and empirical cluster is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}. We will consider a related metric, the volume of the symmetric set difference, which weights points according to their degree in $G_{n,r}$. 
\begin{definition}
	\label{def:volume_symmetric_set_difference}
	For an estimator $\widehat{C} \subseteq \Xbf$ and a set $\mathcal{S} \subseteq \Rd$, the symmetric set difference is 
	\begin{equation*}
	\widehat{C} \vartriangle \mathcal{S}[\Xbf] := \bigl(\widehat{C} \cap \setminus \mathcal{S}[\Xbf]\bigr) \cup \bigl(\Xbf\setminus\widehat{C} \cap \mathcal{S}[\Xbf]\bigr).
	\end{equation*}
	We will measure the discrepancy of $\widehat{C}$ to the set $\mathcal{S}[\Xbf]$ by the volume of the symmetric set difference,
	\begin{equation*}
	\Delta(\widehat{C}, \mathcal{S}[\Xbf]) := \vol(\widehat{C} \vartriangle \mathcal{S}[\Xbf]; G_{n,r})
	\end{equation*}
\end{definition}
However, the symmetric set difference does not measure whether \smash{$\Cest$} 
can distinguish any two distinct clusters \smash{$\Cset,\Cset' \in
\Cbb_f(\lambda)$}. We therefore also study a second notion of cluster
estimation, first introduced by \citet{hartigan1981}, and defined
asymptotically.

\begin{definition}[Consistent density cluster estimation]
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
  \smash{$\Cset \in \Cbb_f(\lambda)$}, we say \smash{$\Cest$} is a consistent 
  estimator of $\Cset$ if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with
  $\Cset \not= \Cset'$, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}
\noindent Consistent cluster recovery roughly ensures that, for a given density threshold $\lambda$, the estimated cluster $\Cest$ contains all
points in a true
density cluster $\Cset \in \Cbb_f(\lambda)$, and simultaneously does not contain any points in any other density cluster $\Cset' \in \Cbb_f(\lambda)$.

With these definitions in place, our broad goal will be to understand the extent to which the PPR algorithm is able to
recover a cluster which either guarantees low symmetric set difference to a true density cluster, or which consistently estimates
a true density cluster.

%\subsection{Our Contributions}

\subsection{Related Work and Our Contributions}
In addition to the background on local spectral
clustering given previously, a few related lines of work are worth
highlighting. In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. The ability of spectral algorithms to perform clustering -- community detection-- in this context is well-understood, dating back to \citet{mcsherry2001} who gives conditions under which the entire community structure can be recovered. In more recent work, \citet{rohe2011} upper bound the fraction of nodes misclassified by a spectral algorithm for the high-dimensional (large number of blocks) SBM and \citet{lei2015} extend these results to the sparse (low average degree) regime. Relatedly, \citet{clauset08,balakrishnan2011,li2018}, analyze the misclassification rate when the block model exhibits some hierarchical structure. The framework we consider, where nodes correspond to data points sampled from an underlying density, and edges between nodes are formed based on geometric proximity, is quite different than the SBM, and therefore so is our analysis.

The study of spectral algorithms on neighborhood graphs has focused on establishing consistency results-- that is, asymptotic convergence of eigenvalues and eigenvectors of matrices to eigenvalues and eigenfunctions of limiting operators. \citet{koltchinskii2000} establish convergence of spectral projections of the adjacency matrix to a limiting integral operator, with similar results obtained using simplified proofs in \citet{rosasco10}. \citet{vonluxburg2008} studies convergence of eigenvectors of the Laplacian matrix for a neighborhood of fixed radius. \citet{belkin07} and \citet{garciatrillos18} extend these results to the regime where the radius $r \to 0$ as $n \to \infty$.

These results are of fundamental importance; however, the behavior of the spectra of these continuum operators can in general be hard to grasp. Therefore, further work relating this spectra to the geometry of the distribution $\Pbb$ is of interest. In this spirit, \citet{shi2009,schiebinger2015,garciatrillos19} examine the ability of  
spectral algorithms to recover the latent labels in certain geometrically well-conditioned nonparametric mixture models. Their results focus on global rather than local 
methods, and thus impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density 
clusters, which is the focus in our work. Perhaps most importantly, these works
rely on general cluster saliency conditions, which implicitly depend on many
distinct geometric aspects of the cluster $\Cset$ under consideration. We make
this dependence more explicit, and in doing so expose the role each geometric
condition plays in the clustering problem.

Our analysis naturally builds on a few of the aforementioned theoretical analyses of PPR. For an arbitrary graph $G$ and subset $S \subset G$, \citet{andersen2006} relate the quality of the PPR cluster $\hat{C}$ to the normalized cut functional $\Phi(S;G)$. While this analysis is tight, it fails to account for improvements possible when the cluster $S$ is additionally assumed to be internally well-connected, which is intuitively a more favorable case for clustering. Building on this intuition, \citet{zhu2013} assume that the subgraph $G[S]$ is internally well-connected-- measured by a functional such as mixing time of a random walk over $G[S]$-- and prove upper bounds on $\vol(\widehat{C} \vartriangle S;G)$. Both of these analyzes hold with respect to a approximate form of PPR (APPR) which can be efficiently computed.

We apply these results to our setting by estimating these functionals in the particular case of $G = G_{n,r}$ and $S = \mathcal{C}[X]$. In particular, we prove an upper bound on the mixing time of a random walk run only over the subset of nodes in $G_{n,r}$ which fall within a density cluster $\Cset$. To do so, we rely on a series of seminal works upper bounding the mixing time of \emph{geometric random walks} (see \citet{vempala2005} for a comprehensive review.)  This study was initiated by \citet{dyer1991}, who used geometric random walks as a fundamental subroutine to efficiently compute volumes of high-dimensional convex bodies. These results are improved in \citet{lovasz1990,kannan97,kannan06}, who show, inter alia, that the bounds on mixing time can be sharpened by avoiding so-called ``start-penalties''. As we discuss further in the sequel, these improvements are crucial to our work. Following the work of \citet{abbasi-yadkori2016,abbasi-yadkori2016a}, we extend these results to hold for Lipschitz deformations of convex sets. Additionally, we relate the mixing time of these (continuous-space) geometric random walks to the mixing time of random walks over (discrete) neighborhood graphs.

Finally, it is worth mentioning that density clustering and level set estimation are well-studied problems. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level-set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010, balakrishnan2013,kpotufe11} consider consistent estimation of the
cluster tree.
%, to note but a few works. 
We emphasize that our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is to start with a 
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

>>>>

\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

@@

\newcommand{\diam}{\rho}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\seq}[1]{\left\{#1\right\}_{n \in \mathbb{N}}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\usepackage{bbm}
\newcommand{\1}{\mathbbm{1}}
\newcommand{\Phibf}{\Phi_{u}}
\newcommand{\Psibf}{\Psi_{u}}
\newcommand{\taubf}{\tau_{u}}
\newcommand{\dist}{\mathrm{dist}}
\newcommand{\Err}{\mathrm{Err}}
\newcommand{\TV}{\mathrm{TV}}

%%% Vectors (no bold font)
\newcommand{\pbf}{p}        % removed bold font
\newcommand{\qbf}{q}
\newcommand{\ebf}[1]{{e}_{#1}}
\newcommand{\pibf}{\pi}

%%% Matrices (no bold font)
\newcommand{\Abf}{A}       % removed bold font 
\newcommand{\Xbf}{X}             
\newcommand{\Wbf}{W}
\newcommand{\Lbf}{L}
\newcommand{\Dbf}{D}
\newcommand{\Ibf}[1]{I_{#1}}

@@

{\large{
\begin{center}
Alden Green~~~~~ Sivaraman Balakrishnan~~~~~ Ryan J. Tibshirani\\
\vspace{.2cm}
\end{center}

@@

\begin{tabular}{c}
\texttt{\{ajgreen,sbalakri,ryantibs\}@stat.cmu.edu}
\end{tabular}
}}

@@

\begin{abstract}
We analyze the Personalized PageRank (PPR) algorithm, a local spectral method
for clustering, which extracts clusters using locally-biased random walks around
a given seed node.  In contrast to previous work, we adopt a classical
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters, that satisfy a set of natural geometric
conditions. We also show a converse result, that PPR can fail to recover
geometrically poorly-conditioned density clusters, even asymptotically. Finally,
we provide empirical support our theory.
\end{abstract}
\end{center}

\section{Introduction}
\label{sec: introduction}

In this paper, we consider the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  Our particular focus is on spectral clustering
methods, a family of powerful nonparametric clustering algorithms. Roughly
speaking, a spectral algorithm first constructs a geometric graph $G$, where
vertices correspond to samples, and edges correspond to proximities between
samples. The algorithm then estimates a feature embedding based on (an
appropriate) Laplacian matrix of $G$, and applies a simple clustering technique
(like $k$-means clustering) in the embedded feature space.

When applied to geometric graphs built from a large number of samples, global
spectral clustering methods can be computationally cumbersome and insensitive to
the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to increased interest in
\emph{local} spectral clustering algorithms, which leverage locally-biased
spectra computed using random walks around some user-specified seed node.  A
popular local clustering algorithm is the Personalized PageRank (PPR) algorithm,
first introduced by \citet{haveliwala2003}, and then further developed by
several others
\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013}.

Local spectral clustering techniques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (such as
conductance).  In this paper, we adopt a classical statistical viewpoint, and
examine what the output of local clustering on a data set reveals about the
underlying density $f$ of the samples.  In particular, we examine the ability of
PPR to recover \emph{density clusters} of $f$, defined as the connected
components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$ for some
$\lambda > 0$ (an object of central interest in the statistical clustering
literature, dating back to the work of \citet{hartigan1981}).

\section{Background and related work}

We begin by providing some standard background on the PPR algorithm
and the density clustering setup, before turning our attention to related work
and a detailed summary of our contributions.

\subsection{PPR on a neighborhood graph}

Let $\Xbf = \{x_1,\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution
$\Pbb$ on $\Rd$, with density $f$.  For a radius $r > 0$, we define
$G_{n,r}=(V,E)$ to be the \emph{$r$-neighborhood graph} of $\Xbf$, an
unweighted, undirected graph with vertices $V=\Xbf$, and an edge $(x_i,x_j) \in
E$ if and only if $\norm{x_i - x_j} \leq r$, where $\norm{\cdot}$ is the
Euclidean norm. We denote by $\Abf \in \Reals^{n \times n}$ the adjacency
matrix, with entries $\Abf_{uv} = 1$ if $(u,v) \in E$ and $0$ otherwise.  We
also denote by $\Dbf$ the diagonal degree matrix, with entries $\Dbf_{uu} :=
\sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n \times n$ identity matrix.

First, we define the PPR vector $\pbf_v = \pbf(v,\alpha;G_{n,r})$, based on 
a given seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to
be the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf_v = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in
position $v$ and 0 elsewhere).

%\subsubsection{Sweep cut}
%\label{subsection: sweep_cut}
Next, we define a \emph{$\beta$-sweep cut} of $\pbf_v = (\pbf_v(u))_{u \in V}$,
for a given level $\beta > 0$, as    
\begin{equation}
\label{eqn: sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{\pbf_v(u)}{\Dbf_{uu}} > \beta}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
then define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over \smash{$\beta \in (L, U)$} (where
the range $(L,U)$ is user-specified), we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} with minimum normalized cut.  For concreteness,
the PPR algorithm is summarized in Algorithm~\ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, sweep cut range $(L,U)$. \\     
{\bfseries Output:} cluster estimate $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf_v=\pbf(v, \alpha; G_{n,r})$ as in
  \eqref{eqn: ppr_vector}.  
  \STATE For \smash{$\beta \in (L,U)$}, compute sweep cuts $S_{\beta}$ as in
  \eqref{eqn: sweep_cuts}. 
  \STATE Return the cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (L,U)}~ \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Estimation of density clusters}

Let \smash{$\Cbb_f(\lambda)$} denote the connected components of the density
upper level set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster
\smash{$\Cset \in \Cbb_f(\lambda)$}, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the
\emph{empirical density cluster}. The size of the symmetric set difference
between estimated and empirical cluster is a commonly used metric to quantify
cluster estimation error \citep{korostelev1993,polonik1995,rigollet2009}. We
will consider a related metric, the volume of the symmetric set difference,
which weights points according to their degree in $G_{n,r}$.
\begin{definition}
  \label{def:volume_symmetric_set_difference}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and a set 
  $\mathcal{S} \subseteq \Rd$, their symmetric set difference is 
  $$
  \Cest \vartriangle \mathcal{S}[\Xbf] :=
    \bigl(\Cest \setminus \mathcal{S}[\Xbf]\bigr) \cup
    \bigl(\mathcal{S}[\Xbf] \setminus \Cest\bigr).
  $$
  Furthermore, we denote the volume of the symmetric set difference by 
  $$
  \Delta(\Cest, \mathcal{S}[\Xbf]) := \vol(\Cest \vartriangle \mathcal{S}[\Xbf];
  G_{n,r}). 
  $$
\end{definition}

Note that the symmetric set difference does not measure whether \smash{$\Cest$}
can (perfectly) distinguish any two distinct clusters \smash{$\Cset,\Cset' \in 
  \Cbb_f(\lambda)$}. We therefore also study a second notion of cluster 
estimation, first introduced by \citet{hartigan1981}, and defined
asymptotically.

\begin{definition}
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster \smash{$\Cset \in
    \Cbb_f(\lambda)$}, we call \smash{$\Cest$} \emph{consistent} for
  \smash{$\Cset$} if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with  
  \smash{$\Cset \not= \Cset'$}, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}

Consistent cluster recovery roughly ensures that, for a given 
threshold $\lambda>0$, the estimated cluster \smash{$\Cest$} contains all points
in a true density cluster $\Cset \in \Cbb_f(\lambda)$, and simultaneously does 
not contain any points in any other density cluster $\Cset' \in
\Cbb_f(\lambda)$.

With these definitions in place, our broad goal will be to understand the extent
to which the PPR algorithm is able to recover a cluster which either guarantees
a low symmetric set difference to a true density cluster, or which consistently
estimates a true density cluster.

\subsection{Related work}

In addition to the background on local spectral clustering given above, a
few related lines of work are worth highlighting. In the stochastic block model
(SBM), arguably one of the simplest models of network formation, edges between
nodes independently occur with probability based on a latent community
membership. In the SBM, the ability of spectral algorithms to perform
clustering---or community detection---is well-understood, dating 
back to \citet{mcsherry2001} who gives conditions under which the entire
community structure can be recovered. In more recent work, \citet{rohe2011}
upper bound the fraction of nodes misclassified by a spectral algorithm for the
high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these
results to the sparse (low average degree) regime. Relatedly,
\citet{clauset08,balakrishnan2011,li2018}, analyze the misclassification rate
when the block model exhibits some hierarchical structure. The framework we
consider, in which nodes correspond to data points sampled from an underlying 
density, and edges between nodes are formed based on geometric proximity, is
quite different than the SBM, and therefore so is our analysis.

In general, the study of spectral algorithms on neighborhood graphs has been
focused on establishing asymptotic convergence of eigenvalues and eigenvectors
of certain sample objects to the eigenvalues and eigenfunctions of corresponding
limiting operators. \citet{koltchinskii2000} establish convergence of spectral
projections of the adjacency matrix to a limiting integral operator, with
similar results obtained using simplified proofs in
\citet{rosasco10}. \citet{vonluxburg2008} studies convergence of eigenvectors of
the Laplacian matrix for a neighborhood of fixed radius. \citet{belkin07} and
\citet{garciatrillos18} extend these results to the regime where the radius $r
\to 0$ as $n \to \infty$.

These results are of fundamental importance; however, the behavior of the
spectra of these continuum operators can in general be hard to grasp. Therefore,
further work relating this spectra to the geometry of the unlerlying
distribution $\Pbb$ is of interest. In this spirit,
\citet{shi2009,schiebinger2015,garciatrillos19} examine the ability of spectral
algorithms to recover the latent labels in certain geometrically
well-conditioned nonparametric mixture models. Their results focus on global
rather than local methods, and thus impose global rather than local conditions
on the nature of the density. Moreover, they do not in general guarantee
recovery of density clusters, which is the focus in our work. Perhaps most
importantly, these works rely on general cluster saliency conditions, which
implicitly depend on many distinct geometric aspects of the cluster $\Cset$
under consideration. We make this dependence more explicit, and in doing so
expose the role each geometric condition plays in the clustering problem.

Our analysis naturally builds on a few of the aforementioned theoretical
analyses of PPR. For an arbitrary graph $G$ and subset $S \subseteq G$,
\citet{andersen2006} relate the quality of the PPR cluster \smash{$\Cest$} to
the normalized cut functional $\Phi(S;G)$. While this analysis is tight in a
worst-case sense, it fails to account for possible improvements when the cluster
$S$ is additionally assumed to be internally well-connected, which is an
intuitively more favorable case for clustering. Building on this intuition,  
\citet{zhu2013} assume that the subgraph $G[S]$ is internally
well-connected---as measured by a functional such as mixing time of a random
walk over $G[S]$---and prove upper bounds on \smash{$\vol(\Cest \vartriangle
  S;G)$}. Both of these analyses also hold with respect to an approximate form 
of PPR  (APPR), which can be efficiently computed.

We apply these results to our setting by carefully analyzing the normalized cut
and mixing time functionals in the particular case of $G = G_{n,r}$ and $S =
\Cset[X]$. One of our main challenges is to prove an upper bound on the mixing 
time of a random walk run only over the subset of nodes in $G_{n,r}$ which fall
within a density cluster $\Cset$. To do so, we rely on a series of seminal works
upper bounding the mixing time of \emph{geometric random walks} (see 
\citet{vempala2005} for a comprehensive review.)  This study was initiated by
\citet{dyer1991}, who used geometric random walks as a fundamental subroutine to
efficiently compute volumes of high-dimensional convex bodies. These results are
improved in \citet{lovasz1990,kannan97,kannan06}, who show, inter alia, that the
bounds on mixing time can be sharpened by avoiding so-called
``start-penalties''. As we discuss further in what follows, these improvements
are crucial to our work. Following the work of
\citet{abbasi-yadkori2016,abbasi-yadkori2016a}, we extend these results to hold
for Lipschitz deformations of convex sets. Additionally, we relate the mixing
time of these (continuous-space) geometric random walks to the mixing time of
random walks over (discrete) neighborhood graphs.

Finally, it is worth mentioning that density clustering and level set estimation
are themselves very well-studied problems in statistics. \citet{polonik1995,
  rigollet2009} study density clustering under the symmetric set difference
metric, \citet{tsybakov1997,singh2009} describe minimax optimal level-set
estimators under Hausdorff loss and
\citet{hartigan1981,chaudhuri2010,balakrishnan2013,kpotufe11} consider
consistent estimation of the cluster tree. We emphasize that our goal is
not to improve on these results, nor to offer a better algorithm for level set
estimation; indeed, seen as a density clustering algorithm, PPR has none 
of the optimality guarantees found in the aforementioned works. Instead, our
motivation is to start with a widely-used local spectral method, PPR, and to
better understand and characterize the distinctions between those density 
clusters which are well-conditioned for PPR, and those which are not.

\subsection{Summary of results}

A summary of our results (and an outline for this paper) is as follows.

\begin{enumerate}
\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we introduce 
  a set of natural geometric conditions on the density cluster $\Cset$
  %formalize a measure of difficulty based on these geometric conditions, 
  and show that if Algorithm~\ref{alg: ppr} is properly initialized, then the size of
  the symmetric set difference between \smash{$\Cest$} and a thickened version
  of the density cluster $\Csig$ can be bounded in a meaningful way based on the
  geometric parameters.

\item We further show in Section \ref{sec:
    consistent_cluster_estimation_with_ppr} that if the density cluster  
  $\Cset$ is particularly well-conditioned, then Algorithm~\ref{alg: ppr} 
  will consistently estimate a density cluster in the strong sense of
  \eqref{eqn: consistent_density_cluster_recovery}.
